---
title: "Untitled"
output: html_document
date: "2025-02-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(assert)
library(vegan)
library(pheatmap)
library(plotrix)
```

```{r}
genus_cpm_df <- read_tsv("S1_filtered_cpm.tsv")
genus_cpm <- as.matrix(genus_cpm_df[, -1])
rownames(genus_cpm) <- genus_cpm_df[[1]]
```

```{r}
meta <- read_tsv("tcga_wgs_primary_tumor_file_meta.tsv", show_col_types = FALSE)
```

Sum the *CPM* values for all genera.

```{r}
genus_count <- rowSums(genus_cpm) %>%
  enframe(name = "genus", value = "count") %>%
  arrange(desc(count)) %>%
  mutate(genus = factor(genus, levels = genus))
```

## Figure 1A
Pie chart of fractions of the top 10 genera, the rest classified as "Other"
```{r}
pie_chart_df <- rbind(
  genus_count[1:10, ],
  tibble(
    genus = "Other",
    count = sum(genus_count$count[11:nrow(genus_count)])
  )
) %>%
  rename(Genus = genus)
figure_1B <- ggplot(
  data = pie_chart_df,
  aes(
    x = "", y = round(100 * count / sum(count), digits = 1),
    fill = Genus
  )
) +
  geom_col() +
  coord_polar(theta = "y") +
  theme_void() +
  theme(
    legend.text = element_text(size = 18),
    legend.title = element_text(size = 20)
  ) +
  xlab("Fraction of Reads per Genus") +
  ylab("Total Normalized Reads") +
  scale_fill_viridis_d(option = "C")
print(figure_1B)
```

Compute mean and standard error of the CPM read counts over all cancer types.
```{r}
avg_project_genera <- double()
sd_project_genera <- double()
total_samples <- double()
for (project in sort(unique(meta$project_id))) {
  these_ids <- meta %>%
    filter(project == project_id) %>%
    pull(file_id)
  this_mat <- genus_cpm[, colnames(genus_cpm) %in% these_ids]
  total_samples[[project]] <- ncol(this_mat)
  avg_project_genera[[project]] <- mean(colSums(this_mat > 0))
  sd_project_genera[[project]] <- std.error(colSums(this_mat > 0))
}
```

Plot the mean and standard error of CPM read counts as a bar chart.
```{r}
mean_genera_df <- tibble(
  project = names(avg_project_genera),
  genera = avg_project_genera,
  sd_genera = sd_project_genera,
  lower = pmax(0, avg_project_genera - sd_project_genera),
  upper = avg_project_genera + sd_project_genera
) %>% arrange(genera)
mean_genera_df$project <- factor(mean_genera_df$project, levels = rev(mean_genera_df$project))
mean_genera_plot <- ggplot(
  data = mean_genera_df,
  aes(x = project, y = genera, fill = project)
) +
  scale_fill_manual(values = rep(c("darkgrey", "lightblue"), 33)) +
  geom_col(show.legend = FALSE) +
  geom_errorbar(aes(
    ymin = lower,
    ymax = upper,
  )) +
  theme_classic() +
  xlab("TCGA Project") +
  ylab("Genera") +
  ggtitle("Mean Genera Per Sample")  +
  theme(axis.text = element_text(size=18), axis.text.x = element_text(size = 12, angle=90, vjust = 0.5, hjust=1),
        axis.title=element_text(size=20),
        plot.title = element_text(size=24))  
print(mean_genera_plot)
```
Pool the CPM normalized sample reads by project.
```{r}
result <- list()
for (project in sort(unique(meta$project_id))) {
  these_ids <- meta %>%
    filter(project == project_id) %>%
    pull(file_id)
  result[[project]] <-
    genus_cpm[, colnames(genus_cpm) %in% these_ids, drop = FALSE] %>% rowSums()
}
project_mat <- do.call(rbind, result)
rownames(project_mat) <- names(result)
```


Calculate and plot the Simpson diversity.
```{r}
simpson_diversity <- diversity(project_mat, index = "simpson") %>%
  sort(decreasing = FALSE)

data <- tibble(
  project = rev(names(simpson_diversity)),
  simpson = rev(simpson_diversity)
)
data$project <- factor(data$project, levels = data$project)
simpson_index_plot <- ggplot(
  data = data,
  aes(x = project, y = simpson, fill = project)
) +
  scale_fill_manual(values = rep(c("darkgrey", "lightblue"), 33)) +
  geom_col(show.legend = FALSE) +
  theme_classic() +
  xlab("TCGA Project") +
  ylab("Simpson Index") +
  ggtitle("Microbiome Diversity Simpson Index") +
  theme(axis.text = element_text(size=18), axis.text.x = element_text(size = 12,  angle=90, vjust = 0.5, hjust=1),
        axis.title=element_text(size=20),
        plot.title = element_text(size=24)) 
print(simpson_index_plot)
```
Plot on a log scale the mean CPM for the top 20 genera across samples in the cancer type.
Cancer types are sorted by mean microbial CPM per sample.
```{r}
abundant_genera = genus_count[1:20,]
mean_cpm = rowSums(project_mat) / total_samples
mean_cpm = sort(mean_cpm, decreasing=TRUE)
pm = project_mat / total_samples
pm = pm[names(mean_cpm),as.character(abundant_genera$genus)]
```

```{r, fig.width=10, fig.height=9}
pheatmap(t(log10(pm+1)), cluster_cols=FALSE,  treeheight_col = 0, treeheight_row = 0,fontsize = 16, cellwidth = 16, angle_col=90)
```

Plotting the Marista-Horn heatmap, though we won't use it directly, but rather just
take the dendrogram.  I cannot find a way to *not* plot the full heatmap.

```{r}
distance <- vegdist(project_mat, "horn")
horn_heatmap = pheatmap(as.matrix(distance),
    clustering_distance_rows = distance,
    clustering_distance_cols = distance, treeheight_col = 0,
    treeheight_row = 0, fontsize = 9,
    main = "Morisita-Horn (abundance) similariy between TCGA projects"
)
```

But now that we have generated the heatmap, we may plot the dendrogram.
```{r, fig.width=9,fig.height=8.1}
library(dendextend)

dendrogram <- horn_heatmap[[1]] %>%
 as.dendrogram() %>%
 set("branches_k_color", k=4) %>%
 set("branches_lwd", 3) %>%
 set("labels_cex", 1.1)

dendrogram %>% plot( horiz=F)
```

Write the figure files.
```{r}
pdf('figure1B.pdf')
print(figure_1B)
dev.off()
pdf('figure2A.pdf')
print(mean_genera_plot)
dev.off()
pdf('figure2B.pdf')
print(simpson_index_plot)
dev.off()
# Just replot the heatmap.
pdf('figure2C.pdf', width=10, height=9)
pheatmap(t(log10(pm+1)), cluster_cols=FALSE,  treeheight_col = 0, treeheight_row = 0,fontsize = 16, cellwidth = 16, angle_col=90)
dev.off()
pdf('figure2D.pdf', width=9, height=8.1)
plot(dendrogram, horiz=F)
dev.off()
```

